Child survival for mothers: mortality change and related measures

### 2020-08-15 paper.Rmd
mods for alap

---
title: 'Child survival for mothers: mortality change and related measures'
author: "Iván Williams and Diego Alburez-Gutierrez"
date: "May 2020"
output:
  html_document:
    fig_caption: yes
bibliography: bib.bibtex
---

@@

$$CS_a \approx l_{a-\kappa} \int_{0}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
(l_{a-\kappa})^{''}\int_{0}^a{\frac{(x-\kappa)^2}{2} m_{x}dx}\\
  \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
  \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]$$

@@

We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function and the survival of daughters from birth to $a-\kappa$. Also that the approximation is affected (negativally) by how disperse is the fertility by age (variance $\sigma^2$) and the survival curvature around (with negative sign), in an age range where $l_x$ is tipycally very flat (range 20-40 years old) in transitioned populations^[This approximation is useful also for get a dimension changes in the fertility average age $\kappa$, keeping the level: taking logs in $F_a \, l_{a-\kappa}$ and deriving we get $\frac{\Delta CS_a}{CS_a} \approx - \mu_{a-\kappa} \Delta\kappa$].

@@

$$\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} x m_{x} {l_{a-x} dx}\\
= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}$$

@@

$$CS_{a}^{\delta_0} = \int_{0}^{a-1} m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + \int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}dx\\
=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx$$

Deriving by $\delta$ and valuating near 0, we get:

@@

This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (cohorts aged more than 1), except for "half" the portion $\frac{CS_{a-1,a}}{CS_a}$ on how important was this las age in total "successful" experience (in terms of alive descendants). If *a* is near *\beta* then $\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx -\Delta\delta$.

@@

$$\frac{d CS_{a}^\delta}{d \delta} = \int_{0}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
= - \int_{0}^{a} {m_{x} l_{a-x} \int_{0}^{a-x}{\mu_t \, dt}\, dx}\\
= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt$$

The last integral is $CS_{a-t}$. Dividing by $CS_a$ and multiplying by $\delta$ get one useful expression. This is the negative of cumulative hazard $H_a$ but considering a positive factor $\frac{CS_{a-t}}{CS_a}$ that takes in account the relative amount of surviving descendants that would be lost at each child age at risk.

@@

Grasping this relationship intuitively is more difficult given the interaction of birth and mortality rates (@KeyfitzMath). 
This factor has a S-shape because $l_x$ curvature and fertility accumulation, that gives more weight to first ages, generally less than 1 except for mother ages close to age valuation *a* with complete reproductive life and high infant mortality (no additional children and high risk for one age additional of survive), typically pretransitional countries with a near 50. A numerical evaluation of both analytical expressions (absolute an relative) are done in [Applications](#Applications) section.

@@

$$MAL_a = \frac{\int_0^a{m_x {\frac{\int_0^{a-x} l_t \, \mu_t (x+t) dt}{\int_0^{a-x} l_t \, \mu_t dt}}dx}}{\int_0^a{m_x}dx} \\
  MAL_a = \frac{\int_0^a{m_x \left[{x+\frac{\int_0^{a-x} l_t \, \mu_t \, t \, dt}{\int_0^{a-x} l_t \, \mu_t dt}}\right]dx}}{\int_0^a{m_x}} \\
  MAL_a = \frac{\int_0^a{m_x \,x \,dx}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a} \\
  MAL_a = \kappa + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a}$$

@@

$$\overline{CS}_a = \int_{0}^{\infty}\int_{0}^{\infty}\left[\int_{0}^{a}{ {m(k)}_x l(z)_{a-x}\,dx}\right]f_{kz}\,dz \,dk\\
= F_\beta \int_{0}^{a-x} r_x \int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z} \, k\, f_{kz} \,dk\,dz\,dx$$

@@

$$\int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z \, k\, f_{kz} \,dk\,dz}\\
\int_{0}^{\infty} \int_{0}^{\infty}{e^{-H_x(Y_0 + Y_1)}e^{Y_0 + Y_2}}\\
L_{Y_0}[1-H_x]L_{Y_1}[-H_x]L_{Y_2}\,dY_0\,dY_1\,dY_2$$

@@

The figure \@ref(fig:CS_abs_app) shows that goodness of fit is decreasing with the change size, given that $\delta$ is assumed near zero. To ilustrate that considering $\delta=0.01$ means at age 50, it means for the absolute change case that $\frac{l_{50}}{e^{\delta\,50}} = l_{50}/$`r round(exp((50*0.01)),1)`, a big change). For the relative case was assumed a constant within each interval $\mu_{x+t}=\mu_x$ for $t$ between 0 and 1, $\int_{x}^{x+1}{\mu_t dt} = \log(l_{x+1})-\log(l_{x})$, and was used this approximation on the empirical data $\frac{\Delta CS_{a}}{CS_a} \approx - \left[\sum_{0}^{a-1}{[\log(l_{x+1})-\log(l_{x})]\, \frac{CS_{a-1-x}}{CS_a}\,}\right] \Delta \delta$.

@@

>>>>

---
title: 'Child survival for mothers: mortality change and related measures'
author: "Iván Williams and Diego Alburez-Gutierrez"
date: "May 2020"
output:
  pdf_document: default
  html_document:
    fig_caption: yes
bibliography: bib.bibtex
---

@@

$$
\begin{aligned}
CS_a \approx & l_{a-\kappa} \int_{0}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
(l_{a-\kappa})^{''}\int_{0}^a{\frac{(x-\kappa)^2}{2} m_{x}dx}\\
  & \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
  & \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]
\end{aligned}
$$

@@

We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function and the survival of daughters from birth to $a-\kappa$. The approximation is affected negativally by how disperse is the fertility by age (variance $\sigma^2$) and the survival curvature around (with negative sign), in an age range where $l_x$ is tipycally very flat in transitioned populations (range 20-40 years old)^[This approximation is useful also for get an idea of impact because of changes in the fertility average age $\kappa$, keeping constant the level: taking logs in $F_a \, l_{a-\kappa}$ and deriving we get $\frac{\Delta CS_a}{CS_a} \approx \mu_{a-\kappa} \Delta\kappa$, being daughters exposed to less time beacuse of the delay in pattern].

@@

$$
\begin{aligned}
\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} x m_{x} {l_{a-x} dx}\\
= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}
\end{aligned}
$$

@@

$$
\begin{aligned}
CS_{a}^{\delta_0} = \int_{0}^{a-1} m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + \int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}dx \\
=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx
\end{aligned}
$$
Deriving by $\delta$ and valuating near 0, we get:

@@

This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (cohorts aged more than 1), except for "half" the portion $\frac{CS_{a-1,a}}{CS_a}$ on how important was this las age in total "successful" experience (in terms of alive descendants). If $a$ is near $\beta$ then $\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx -\Delta\delta$.

@@

$$
\begin{aligned}
\frac{d CS_{a}^\delta}{d \delta} = \int_{0}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
= - \int_{0}^{a} {m_{x} l_{a-x} \int_{0}^{a-x}{\mu_t \, dt}\, dx}\\
= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt
\end{aligned}
$$

The last integral is $CS_{a-t}$. Dividing by $CS_a$ and multiplying by $\delta$ get one useful expression. This is the negative of cumulative hazard $H_a$ but considering a positive factor $\frac{CS_{a-t}}{CS_a}$ that takes in account the relative amount of surviving descendants that would be lost at each child age at risk, that gives more weight to first ages because all the parity was exposed in that age.

@@

Grasping this relationship intuitively is more difficult given the interaction of birth and mortality rates (@KeyfitzMath). 
The factor $\frac{CS_{a-t}}{CS_a}$  has a S-shape because $l_x$ curvature and fertility accumulation. Taking the difference between consecutive ages $a$ and $a+1$, we have $CS_{a-1}=\int_{0}^{a-1} {m_{x} l_{a-1} \, dx}$, $CS_{a}=\int_{0}^{a-1} {m_{x} l_{a} \, dx}+CS_{a-1,a}$. In the last one, a second term is added because of additional born daughters after $a-1$, and in the first term the existing one are survived at on $\frac{l_{a-x}}{l_{a-1-x}}$. In advanced ages when fertility is complete, the difference is only explained by the surviving part, that is why this factor is not a strictly increasing function but for mother ages close to $\beta$ (last theoretical fecundity age). A numerical evaluation of both analytical expressions (absolute an relative) and its factor $\frac{CS_{a-t}}{CS_a}$ are done in [Applications](#Applications) section.

@@

$$
\begin{aligned}
MAL_a = & \frac{\int_0^a{m_x {\frac{\int_0^{a-x} l_t \, \mu_t (x+t) dt}{\int_0^{a-x} l_t \, \mu_t dt}}dx}}{\int_0^a{m_x}dx} \\
  MAL_a = & \frac{\int_0^a{m_x \left[{x+\frac{\int_0^{a-x} l_t \, \mu_t \, t \, dt}{\int_0^{a-x} l_t \, \mu_t dt}}\right]dx}}{\int_0^a{m_x}} \\
  MAL_a = & \frac{\int_0^a{m_x \,x \,dx}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a} \\
  MAL_a = & \kappa + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a}
\end{aligned}
$$

@@

$$
\begin{aligned}
\overline{CS}_a = \int_{0}^{\infty}\int_{0}^{\infty}\left[\int_{0}^{a}{ {m(k)}_x l(z)_{a-x}\,dx}\right]f_{kz}\,dz \,dk\\
= F_\beta \int_{0}^{a-x} r_x \int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z} \, k\, f_{kz} \,dk\,dz\,dx
\end{aligned}
$$

@@

$$
\begin{aligned}
\int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z \, k\, f_{kz} \,dk\,dz}\\
\int_{0}^{\infty} \int_{0}^{\infty}{e^{-H_x(Y_0 + Y_1)}e^{Y_0 + Y_2}}\\
L_{Y_0}[1-H_x]L_{Y_1}[-H_x]L_{Y_2}\,dY_0\,dY_1\,dY_2
\end{aligned}
$$

@@

The figure \@ref(fig:CS_abs_app) shows that goodness of fit is decreasing with the change size, given that $\delta$ is assumed near zero. To ilustrate that considering $\delta=0.01$ means at age 50, it means for the absolute change case that $\frac{l_{50}}{e^{\delta\,50}} = l_{50}/$ `r round(exp((50*0.01)),1)`, a big change). For the relative case was assumed a constant within each interval $\mu_{x+t}=\mu_x$ for $t$ between 0 and 1, $\int_{x}^{x+1}{\mu_t dt} = \log(l_{x+1})-\log(l_{x})$, and was used this approximation on the empirical data $\frac{\Delta CS_{a}}{CS_a} \approx - \left[\sum_{0}^{a-1}{[\log(l_{x+1})-\log(l_{x})]\, \frac{CS_{a-1-x}}{CS_a}\,}\right] \Delta \delta$.

@@

*More develop here showing the factor and ggplot type...*

```{r chek_CS_ratio, include = F, eval=F}
# test factor
countries = c("Guatemala", "Argentina", "Haiti")
period = "1950-1955"
Age = 50
plot(NULL, xlim=c(15,Age), ylim=c(0,1.1),xlab="Age", ylab="Ratio")
abline(h=1, lty=2, col="grey")
for (country in countries){
  points(15:Age,
         sapply(15:Age, function(age)
         CS_results(country = country, 
             period = period, 
             a = age)$CS)/
          CS_results(country = country, 
             period = period, 
             a = Age)$CS,
          col = which(country==countries), t="l")
}
```
